Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-16 paper/paper.tex
finished discussion on analysis of existing sequences

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original site of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_without_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_without_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_without_deleted_seqs.pdf}
 }
 \caption{
 Phylogenetic trees of all SARS-CoV-2 sequences with multiple observations among viruses collected no later than January 31, 2020.
 The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
The nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
The mutations on each branch are labeled, with mutations towards the nucleotide identity in the RaTG13 outgroup in purple.
The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as the mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the widely used Wuhan-Hu-1 reference sequence.
All sequences in the monophyletic group that contains C28144T are collapsed into a single clade labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; note that this clade contains Wuhan-Hu-1.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is instead RpYN06 or RmYN02.
\label{fig:tree_RaTG13}
 }
 \end{figure*}

@@

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the three possible locations closest to the bat outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

Before moving onto the next section, I will also briefly address several less plausible explanations for the discordance that have gained traction in popular discussion of SARS-CoV-2's origins.
The first explanation suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
This explanation can be discounted, because although there are some unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the fundamental conundrum about the progenitor described above holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before the emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.

Tables 6 and 7 of WHO-China report indicate viruses from Huanan Market all in clade B except S13, which has T8782, T28144, and some mutation at 4946. However, clades are connected by single mutations, for instance Sichuan/SC-MY-054/2020 has T8782 but C28144.

>>>>

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_without_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_without_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_without_deleted_seqs.pdf}
 }
 \caption{
 Phylogenetic trees of all SARS-CoV-2 sequences with multiple observations among viruses collected no later than January 31, 2020.
 The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
The mutations on each branch are labeled, with mutations towards the nucleotide identity in the RaTG13 outgroup in purple.
The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the Wuhan-Hu-1 reference sequence.
All sequences in the monophyletic group containing C28144T are collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is instead RpYN06 or RmYN02.
\label{fig:tree_RaTG13}
 }
 \end{figure*}

@@

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the bat outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between the phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
This explanation can be discounted, because although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other unknown market being the source of viruses that lack the T8782C and C28144T mutations.
However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows why this idea can be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors into the other after it had jumped to humans.
